Storm surges in the Strait of Georgia simulated with a regional model

### 2015-01-06 AO_guides/manuscriptAO.tex
Minor typographical edits

\author{Nancy Soontiens$^{1}$$^{\ast}$\thanks{$^\ast$Corresponding author's email: nsoontie@eos.ubc.ca
\vspace{6pt}}, Susan Allen$^{1}$,  Doug Latornell$^{1}$,  Kate Le Sou\"{e}f$^{1}$, Idalia Machuca$^{1}$,\\ Jean-Philippe Paquin$^{2}$,
Youyu Lu$^{3}$, Keith Thompson$^{2}$, Vasily Korabel$^{2}$
\\\vspace{6pt}  $^{1}${\em{University of British Columbia}};
$^{2}${\em{Dalhousie University}};
$^{3}${\em{Bedford Insitute of Oceanography}}\\\received{v1.2 released April 2014} }

@@

More recently, modelling studies have focused on baroclinic features. \citet{masson2004modelling} used the Princeton Ocean Model (POM) in a regional mode to examine the seasonal variability of deep water properties in the Strait of Georgia and noted a strong sensitivity to vertical mixing over the sills. This model employed a sigma or terrain-following vertical coordinate system leading to spurious diapycnal mixing since, numerically, the lateral mixing is computed on sigma levels which may intersect isopycnals in the presence of steep topography. 
%A finite volume model on an unstructured grid by \citet{yang2010multi} involved a truncated domain with an open boundary in the northern Strait of Georgia, where under predictions in the M$_2$ tidal phase over most of their domain were suggested to be caused by an inaccurate representation of the tides at their northern boundary. 
\citet{sutherland2011model} employed the Regional Ocean Modelling System (ROMS) to study estuarine circulation in the Salish Sea with a model domain that extended outside the Strait of Juan de Fuca into the Pacific Ocean and up to the northern Strait of Georgia. They noted that their semi-diurnal amplitudes were generally too low. An updated version of this model produced high temperatures in the Strait of Juan de Fuca, suggesting too little mixing within the Salish Sea \citep{giddings2014hindcasts}. Smaller domains have also been considered; notably, the passageways through the Discovery Islands north of the Strait of Georgia have been modelled baroclinically by \citet{foreman2012circulation} with the Finite Volume Coastal Ocean Model (FVCOM) pointing to difficulties in reproducing baroclinic currents on their grid.  Accurate reproduction of observed tidal amplitudes and phases and appropriate treatment of mixing is a recurring difficulty in modelling this region.

@@

We have used the Nucleus for European Modelling of the Ocean (NEMO) framework in its regional configuration \citep{madec2012nemo} to develop an ocean model for the Strait of Georgia and Salish Sea. NEMO is a highly modularized tool used for studying ocean physics, ocean-ice interactions, and the bio-geochemical properties of the ocean. NEMO's ocean core solves the three-dimensional hydrostatic equations of motion for an incompressible fluid under the Boussinesq approximation on a structured computational grid. Although not used in the present work, NEMO's options for grid nesting and bio-geochemical coupling make it a useful tool for studying the complex physics and bio-geochemical interactions within the Strait of Georgia. This work focuses on validating the physical configuration of the Salish Sea model, in particular, determining appropriate forcing and boundary conditions for accurate reproduction of tidal amplitudes and phases as well as storm surge elevations. Future work will include bio-geochemical coupling and data assimilation. NEMO version 3.4 was used in this study.

@@

The ocean surface is forced with momentum and heat fluxes from a 33-km global atmospheric reforecasting model suitable for use in ocean modelling \citep{smith2014new}. Forecasts from the period of 2002-2012 are available. One simulation employs the western Canada component of the High Resolution Deterministic Prediction System (HRDPS), a nested 2.5 km resolution atmospheric model provided by Environment Canada \citep{ECModel}. In addition to momentum and heat fluxes, forcing due to the local atmospheric pressure is also included by calculating an inverse barometer sea surface height,
\begin{equation}
 \eta_{IB} = -\frac{1}{g\rho_{0}}\left(P_{s}-P_0\right), \label{eq:inverse}
\end{equation}
where $g$ is the accelaration due to gravity, $\rho_{0}$ is a reference density set to 1035 kg m$^{-3}$, $P_{s}$ is the sea level atmospheric pressure from the atmospheric model, and $P_0$ is a reference sea level pressure set to 101,000 Pa.  The gradient of $\eta_{IB}$  is scaled by $g$ and added to the forcing terms in the ocean momentum equations \citep{madec2012nemo}. Since the atmospheric model employs a terrain following vertical coordinate the model surface pressure has been corrected to sea level using \citep{holton1992introduction}
\[ P_s = P_1\left(\gamma\frac{z_1}{T_1} +1 \right)^\frac{g}{\gamma R},\]
where $P_1$ and $T_1$ are the model pressure and temperature at height $z_1$, $R$ is the ideal gas constant, and $\gamma$ is a temperature lapse rate set to 0.0098 K m$^{-1}$. Since the model domain does not include the open Pacific Ocean, pressure gradients between the open ocean and the semi-enclosed Salish Sea and their effect in the sea surface height cannot be accounted for. Previous studies in the Hudson Bay and Laborador Shelf system have aruged that oscillations in the sea surface height can be expained, in part, by the response to the pressure gradient between these regions \citep{wright1987influence,young1995synoptic}.

@@

In general, the K$_1$ tidal harmonics are well represented by the model with complex differences below 5 cm at all stations except Seymour Narrows (station 27). Seymour Narrows is a very constricted region in Johnstone Strait and is not well resolved by the model. As such, the M$_2$ amplitude and phase are also poorly reproduced here. The model's M$_2$ amphidrome, indicated by the drop in amplitude around station 5 (Victoria), is too strong resulting in lower than observed M$_2$ amplitudes in this region. Additionally, the M$_2$ phase errors in the Strait of Juan de Fuca (stations 1-7) are quite large, leading to large complex differences in this region.  However, the match in the Strait of Georgia (stations 10-24) is very good, with complex differences less than 6 cm, in part due to tidal tuning for highly accurate tides in the main region of interest. Details on the tidal tuning procedure are provided in Appendix \ref{sec:appendix}.

@@

\section{Storm Surge Hindcasts}\label{sec:storm}
A series of five storm surge hindcasts is presented next. Storm surge simulations were started from rest at least three days prior to the event of interest to allow for adjustment of model currents. Stratification conditions from model spin-up at a time close to the event (within five days) were chosen to initialize the temperature and salinity fields. Boundary conditions include atmospheric forcing and fresh water input from rivers at the surface, sea surface height anomaly from observations at the open boundary, and tidal forcing with eight tidal constituents. Forcing the model with only eight tidal constituents leads to an error in modelled sea surface height due to the omission of the next leading order tidal constituents such as J$_1$. Comparisons between full tidal predictions and tidal predictions using eight constituents suggest that this error can approach 40 cm during times of high tidal range, and in particular, during the storm surge season Nov-Feb. As such, storm surge model predictions were corrected by accounting for this error; for each storm surge location of intetest, the difference between a full tidal prediction and a tidal prediction with eight constiuents is added to the modelled sea surface height. This correction is only applied to the modelled water level and not the modelled residuals. Additionally, since the modelled sea surface height is calculated about a zero reference level, the modelled water level is determined by adding a long term mean sea level at the specified location to the modelled sea surface height. All times are reported in UTC.

@@

The model's skill is reported through a few statistical measures of the the total water level $\eta$ and the residual $\delta$. First, the mean absolute error ($\bar{e}$), defined as
\begin{equation}
\bar{e} = \text{mean}\left(\left| M - O \right|\right),
\end{equation}
where $M$ and $O$ are the modelled and observed quantities of interest, is a common measure that estimates the average error in the model. Next, a measure comparing the variance of the error to the variance of the observations defined as \citep{thompson2003prediction},
\begin{equation}
\gamma^2 = \frac{\text{var}\left(M-O\right)}{\text{var}\left(O\right)},
\end{equation}
is also calculated. Smaller values of $\gamma^2$ indicate better model skill. This quantity was used by \citet{bernier2006predicting} and \citet{bernier2010tide} in storm surge predictions in Eastern Canada and is calculated here as a point of comparison. A third measure, called the Willmott skill score \citep{willmott1982some}, is also used to assess overall model skill in predicting the water level and residual. This quantity is defined as
\begin{equation}
WS = 1 - \frac{\sum_{i=1}^N \left(M_i - O_i\right)^2}{\sum_{i=1}^N \left(|M_i'| + |O_i'|\right)^2},
\end{equation} 
where $M_i' = M_i-\bar{O}$ and $O_i'=O_i-\bar{O}$ and $\bar{O}$ is the mean of the observations. This score is bounded by zero and on and values close to one indicate good model skill. Since the Willmott skill score is nondimensional, it is an approriate statistic to report when comparing skills of different variables, in this case, the total water level and the residuals. Since the total water level has a larger variability than the residuals the $\gamma^2$ measure should be much lower for the total water levels. These three statistics, along with the mean observed values and mean modelled values (which indicate the model's bias) are reported in Table \ref{tab:statistics} for each hindcast.

@@

%Local winds
The contribution to the surge due to atmospheric forcing is small and appears to be geographically dependent, most likely due to the effects of the local winds. Removal of the atmosperhic forcing leads to a drop of about 2.5 cm in the maximum surge at Point Atkinson and 2.3 cm at Campbell River. By contrast, the surge amplitude increases by 4.8 cm at Victoria and 3.6 cm at Patricia Bay when atmospheric forces are neglected. The southeasterly winds over the Strait of Georgia could be acting to push water away from the southern tip of Vancouver Island, resulting in a slight decrease in the surge amplitude at Victoria and Patricia Bay when wind effects are included in the model. A simulation with only wind forcing produces a surge of 7.8 cm at Point Atkinson, a small fraction of the total observed surge, and only centimetres at the other locations.  Note that this wind-induced surge occurs well after the peak surge occurred in both the observations and simulations with full forcing.

@@

%local pressure
Next, the surge amplitude decreases by a few centimetres at all four locations if local atmospheric pressure is not included in the surface forcing. Other findings have reported the inverse barometer effect due to atmospheric pressure contributes up to $^1/_3$ of the total surge amplitude \citep{murty1995storm}. Note that in these simulations the removal of the atmospheric pressure in the surface forcing does not entirely remove the inverese barometer effect from the model results since the sea surface height anomaly at the open boundary includes an inverse barometer component. As such, these simulations indicate the importance of the local atmospheric pressure forcing but not the inverse barometer entirely. The contribution due to the inverse barometer can be calculated with equation \ref{eq:inverse} or by the approximation that 1 hPa drop in pressure corresponds to 1 cm rise in sea level. In this event, the pressure at the Vancouver International Airport dropped to about 99 000 Pa \citep{ECClimateArchive}, corresponding to a 20 cm rise in sea level. This inverse barometer contribution is more inline with the findings of \citep{murty1995storm} but does not alter the conclusion that the remote forcing contributes most significantly. Further, the model calculates the local atmopsheric pressure forcing using a spatial gradient, and since the spatial scale of the low pressure system is large compared to the domain the local atmospheric pressure should not contribute greatly.

@@

Next, an assessment of the importance of including the tide-surge interaction, which is important in shallow regions due to nonlinear effects and bottom friction \citep{bernier2007tide}, is presented for this hindcast (Figure \ref{fig:tidesurge}). The most significant differences occur at Point Atkinson where the maximum of the surge-only case is 8.7 cm higher than maximum modelled residual, about 10\% of the total surge amplitude.  In contrast, the tide-surge interaction at Victoria is hardly noticeable. Indeed, a spatial map of the difference between the modelled residual and surge-only difference (Figure \ref{fig:tidesurge_spatial}) indicates that the tide-surge interaction is more pronounced north of the San Juan and Gulf Islands. The passages between these islands are known for vigorous tidal mixing, which may have a nonlinear effect on the surge propagation. However, the effect is a relatively small overestimate of the surge elevation and no changes to the timing of the surge. So, in operational settings, performing a surge-only simulation and adding the predicted tides \textit{a posteriori} could be more computationally efficient with only a small decrease in the accuracy of the prediction.

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/feb2006_tidesurge_spatial.pdf}
\caption{Spatial dependency of tide-surge interaction Difference in sea surface height between simulations with tidal forcing and without tidal forcing. Each plot is an average over 4 hours. }
\label{fig:tidesurge_spatial}
\end{figure}

@@

Modelled residuals are generally too low in this hindcast (Figure \ref{fig:dec2006}), however, the timing of the surge agrees well. All Willmott skill scores rank above 0.88 and the $\gamma^2$ measure also indicates good agreement between model and observations (Table \ref{tab:statistics}). The $\gamma^2$ measure suggests that the standard deviation of the error in the modelled resdiuals is roughly half of the standard deviation of the observed residuals. Note that the performance has degraded when compared to the Feb 2006 and Nov 2006 cases. This could be due to a number of factors. First, the very large wind speeds observed during this storm are significantly underrepresented by the atmospheric model leading to lower wind stress and less piling of water against the coast. Second, the anomaly forcing at the open boundaries could be too low. The observed residual at Tofino, BC (a point near by but outside of the domain) is up to 10 cm lower than the observed residual at Neah Bay, WA (a point near by but inside of the domain). Since the storm surge is less sensitive to wind forcing at the surface than anomaly forcing at the open boundaries, the latter is likely the cause.

@@

Hindcasts with both atmospheric forcing products agree well with the observations (Figure \ref{fig:dec2012_surge}). Willmott skill scores for both modelled residuals are above 0.92 at these four locations. Further, the modelled residuals deviate only slightly from each other, most noticeable at Point Atkinson during the large wind event on Dec 15, 2012 where simulation with the low resolution atmospheric forcing has a residual that is 5 cm higher. Although the wind speeds on Dec 15, 2012 are quite high, this event did not result in any flooding or residuals greater than 50 cm. Significant flooding occurred later, on Dec 17, 2012, where the observed surge amplitude at Point Atkinson was 68.3 cm. Even though the low resolution model significantly under predicts the wind speed at some locations, it performs comparably to forcing with high resolution products. Much of the fine scale detail offered by the high resolution atmopsheric model is averaged out since, to first order, the storm surge is a spatial integral of the wind. This result further demonstrates that the local wind is not the most important factor for setting up storm surges over this domain.

@@

\section{Conclusion and discussion}\label{sec:diss}
In this article, we have configured a three-dimensional numerical ocean model of the Salish Sea for accurate hindcasts of storm surges in the Strait of Georgia. The model is forced with eight tidal constituents at the western and northern open boundaries, as well as sea surface height anomaly from de-tided observations at Tofino and Port Hardy. Additionally, wind stress and an inverse barometer effect were forced at the ocean surface using an atmospheric reforecasting product called CGRF \citep{smith2014new}. Storm surge simulations began with zero initial conditions for the currents and sea surface height and were commenced several days prior to the event of interest to spin up the velocity fields.

@@

Several factors considered in other storm surge modelling studies have been neglected in these simulations. Primarily, effects due to wetting and drying have not been included. This is an important factor to consider in shallow, low lying areas where water levels retreat significantly to reveal land. In this article, the four locations used to assess storm surge accuracy are not in shallow regions and so wetting and drying is not considered to be important. However, several communities in the Strait of Georgia, such as Boundary Bay and Crescent Beach, are in shallow regions and are typically most affected by storm surge during the winter season. As such, efforts to include wetting and drying will be considered in the future. Lastly, flooding effects due to wave run up is left for a more detailed inundation study with higher resolution in the near-shore areas. Local winds have a significant impact on wave action which is a primary concern for flooding management. Effects due to stratification will also be considered in future work. Further, the remote forcing is applied uniformly across the open boundary and so any spatial differences in the incoming surge are not accounted for in the model.

@@

Finally, difficulties in reproducing accurate M$_2$ tidal harmonics in both the Strait of Georgia and the Strait of Juan de Fuca have been highlighted in this study among others \citep{stronach1993update, foreman2004m}. Future work will attempt to improve model accuracy by examining mixing parametrizations and energy dissipation due to bottom friction and their effect on tidal harmonics. Additionally, a biological-chemical modelling component will be added. Stratification effects on storm surge will be considered in a future study to determine if a barotropic model is reasonably accurate.

@@

\appendices
\section{Appendix}\label{sec:appendix}
This section details the procedure used to tune the tidal forcing for a good match with observations in the Strait of Georgia. Originally, the eight most important tidal constituents were extracted from a tide prediction tool called Webtide \citep{webtide} at the western Juan de Fuca boundary based on work by \citet{foreman2000webtide}. At the northern boundary,  M$_2$ and K$_1$ tidal harmonics for both elevation and currents were taken from \citet{thomson1980johnstone}, as well as elevation harmonics for O$_1$ and S$_2$. The remaining constituents were extrapolated from Webtide based on the ratio between S$_2$/M$_2$ amplitudes and phase differences for the semi-diurnal constituents and O$_1$/K$_1$ for the diurnal constituents. However, these forcing conditions did not result in a good match between observed and modelled tidal elevation harmonics in the Strait of Georgia, the main region of interest (although the matches in the Strait of Juan de Fuca were adequate). As such, tidal forcing was tuned, by first adjusting the western boundary M$_2$ and K$_1$ phase and K$_1$ amplitude for a good match in the Strait of Georgia (specifically at Point Atkinson, Gibsons Landing, Winchelsea and Halfmoon Bay). This tuning was performed on simulations that used only M$_2$ and K$_1$ forcing. Next, the M$_2$ and K$_1$ fluxes at the northern boundary were adjusted because the cross-sectional area of transect used by \citet{thomson1980johnstone} to measure tidal currents in Johnstone Strait is smaller than the cross-sectional area of our northern boundary. Finally, the remaining constituents were added, starting with O$_1$, then S$_2$, and then in pairs P$_1$/N$_2$ and Q$_1$/K$_2$. Final adjustments were made by slightly varying amplitude and phase, evaluating the changes to M$_2$, K$_1$ and the ratios S$_2$/M$_2$, N$_2$/M$_2$, K$_2$/M$_2$, O$_1$/K$_1$, P$_1$/K$_1$, Q$_1$/K$_1$ and then choosing new values to move closer to observations by linear interpolation.

>>>>

\author{Nancy Soontiens$^{1}$$^{\ast}$\thanks{$^\ast$Corresponding author's email: nsoontie@eos.ubc.ca
\vspace{6pt}}, Susan Allen$^{1}$,  Doug Latornell$^{1}$,  Kate Le Sou\"{e}f$^{1}$, Idalia Machuca$^{1}$,\\ Jean-Philippe Paquin$^{2}$,
Youyu Lu$^{3}$, Keith Thompson$^{2}$, Vasily Korabel$^{2}$
\\\vspace{6pt}  $^{1}${\em{University of British Columbia}};
$^{2}${\em{Dalhousie University}};
$^{3}${\em{Bedford Institute of Oceanography}}\\\received{v1.2 released April 2014} }

@@

More recently, modelling studies have focused on baroclinic features. \citet{masson2004modelling} used the Princeton Ocean Model (POM) in a regional mode to examine the seasonal variability of deep water properties in the Strait of Georgia and noted a strong sensitivity to vertical mixing over the sills. This model employed a sigma or terrain-following vertical coordinate system leading to spurious diapycnal mixing since, numerically, the lateral mixing is computed on sigma levels which may intersect isopycnals in the presence of steep topography. 
%A finite volume model on an unstructured grid by \citet{yang2010multi} involved a truncated domain with an open boundary in the northern Strait of Georgia, where under predictions in the M$_2$ tidal phase over most of their domain were suggested to be caused by an inaccurate representation of the tides at their northern boundary. 
\citet{sutherland2011model} employed the Regional Ocean Modelling System (ROMS) to study estuarine circulation in the Salish Sea with a model domain that extended outside the Strait of Juan de Fuca into the Pacific Ocean and up to the northern Strait of Georgia. They noted that their semi-diurnal amplitudes were generally too low. An updated version of this model produced high surface temperatures in the Strait of Juan de Fuca, suggesting too little mixing within the Salish Sea \citep{giddings2014hindcasts}. Smaller domains have also been considered; notably, the passageways through the Discovery Islands north of the Strait of Georgia have been modelled baroclinically by \citet{foreman2012circulation} with the Finite Volume Coastal Ocean Model (FVCOM) pointing to difficulties in reproducing baroclinic currents on their grid.  Accurate reproduction of observed tidal amplitudes and phases and appropriate treatment of mixing is a recurring difficulty in modelling this region.

@@

We have used the Nucleus for European Modelling of the Ocean (NEMO) framework in its regional configuration \citep{madec2012nemo} to develop an ocean model for the Salish Sea. NEMO is a highly modularized tool used for studying ocean physics, ocean-ice interactions, and the bio-geochemical properties of the ocean. NEMO's ocean core solves the three-dimensional hydrostatic equations of motion for an incompressible fluid under the Boussinesq approximation on a structured computational grid. Although not used in the present work, NEMO's options for grid nesting and bio-geochemical coupling make it a useful tool for studying the complex physics and bio-geochemical interactions within the Strait of Georgia. This work focuses on validating the physical configuration of the Salish Sea model, in particular, determining appropriate forcing and boundary conditions for accurate reproduction of tidal amplitudes and phases as well as storm surge elevations. Future work will include bio-geochemical coupling and data assimilation. NEMO version 3.4 was used in this study.

@@

The ocean surface is forced with momentum and heat fluxes from a 33-km global atmospheric reforecasting model suitable for use in ocean modelling \citep{smith2014new}. Forecasts from the period of 2002-2012 are available. One simulation employs the western Canada component of the High Resolution Deterministic Prediction System (HRDPS), a nested 2.5 km resolution atmospheric model provided by Environment Canada \citep{ECModel}. In addition to momentum and heat fluxes, forcing due to the local atmospheric pressure is also included by calculating an inverse barometer sea surface height,
\begin{equation}
 \eta_{IB} = -\frac{1}{g\rho_{0}}\left(P_{s}-P_0\right), \label{eq:inverse}
\end{equation}
where $g$ is the acceleration due to gravity, $\rho_{0}$ is a reference density set to 1035 kg m$^{-3}$, $P_{s}$ is the sea level atmospheric pressure from the atmospheric model, and $P_0$ is a reference sea level pressure set to 101,000 Pa.  The gradient of $\eta_{IB}$  is scaled by $g$ and added to the forcing terms in the ocean momentum equations \citep{madec2012nemo}. Since the atmospheric model employs a terrain following vertical coordinate the model surface pressure has been corrected to sea level using \citep{holton1992introduction}
\[ P_s = P_1\left(\gamma\frac{z_1}{T_1} +1 \right)^\frac{g}{\gamma R},\]
where $P_1$ and $T_1$ are the model pressure and temperature at height $z_1$, $R$ is the ideal gas constant, and $\gamma$ is a temperature lapse rate set to 0.0098 K m$^{-1}$. Since the model domain does not include the open Pacific Ocean, pressure gradients between the open ocean and the semi-enclosed Salish Sea and their effect on the sea surface height cannot be accounted for. Previous studies in the Hudson Bay and Labrador Shelf system have argued that oscillations in the sea surface height can be explained, in part, by the response to the pressure gradient between these regions \citep{wright1987influence,young1995synoptic}.

@@

In general, the K$_1$ tidal harmonics are well-represented by the model with complex differences below 5 cm at all stations except Seymour Narrows (station 27). Seymour Narrows is a very constricted region in Johnstone Strait and is not well resolved by the model. As such, the M$_2$ amplitude and phase are also poorly reproduced here. The model's M$_2$ amphidrome, indicated by the drop in amplitude around station 5 (Victoria), is too strong resulting in lower than observed M$_2$ amplitudes in this region. Additionally, the M$_2$ phase errors in the Strait of Juan de Fuca (stations 1-7) are quite large, leading to large complex differences in this region.  However, the match in the Strait of Georgia (stations 10-24) is very good, with complex differences less than 6 cm, in part due to tidal tuning for highly accurate tides in the main region of interest. Details on the tidal tuning procedure are provided in Appendix \ref{sec:appendix}.

@@

\section{Storm Surge Hindcasts}\label{sec:storm}
A series of five storm surge hindcasts is presented next. Storm surge simulations were started from rest at least three days prior to the event of interest to allow for adjustment of model currents. Stratification conditions from model spin up at a time close to the event (within five days) were chosen to initialize the temperature and salinity fields. Boundary conditions include atmospheric forcing and fresh water input from rivers at the surface, sea surface height anomaly from observations at the open boundary, and tidal forcing with eight tidal constituents. Forcing the model with only eight tidal constituents leads to an error in modelled sea surface height due to the omission of the next leading order tidal constituents such as J$_1$. Comparisons between full tidal predictions and tidal predictions using eight constituents suggest that this error can approach 40 cm during times of high tidal range, and in particular, during the storm surge season Nov-Feb. As such, storm surge model predictions were corrected by accounting for this error; for each storm surge location of interest, the difference between a full tidal prediction and a tidal prediction with eight constituents is added to the modelled sea surface height. This correction is only applied to the modelled water level and not the modelled residuals. Additionally, since the modelled sea surface height is calculated about a zero reference level, the modelled water level is determined by adding a long term mean sea level at the specified location to the modelled sea surface height. All times are reported in UTC.

@@

The model's skill is reported through a few statistical measures of the the total water level $\eta$ and the residual $\delta$. First, the mean absolute error ($\bar{e}$), defined as
\begin{equation}
\bar{e} = \text{mean}\left(\left| M - O \right|\right),
\end{equation}
where $M$ and $O$ are the modelled and observed quantities of interest, is a common measure that estimates the average error in the model. Next, a measure comparing the variance of the error to the variance of the observations defined as \citep{thompson2003prediction},
\begin{equation}
\gamma^2 = \frac{\text{var}\left(M-O\right)}{\text{var}\left(O\right)},
\end{equation}
is also calculated. Smaller values of $\gamma^2$ indicate better model skill. This quantity was used by \citet{bernier2006predicting} and \citet{bernier2010tide} in storm surge predictions in Eastern Canada and is calculated here as a point of comparison. A third measure, called the Willmott skill score \citep{willmott1982some}, is also used to assess overall model skill in predicting the water level and residual. This quantity is defined as
\begin{equation}
WS = 1 - \frac{\sum_{i=1}^N \left(M_i - O_i\right)^2}{\sum_{i=1}^N \left(|M_i'| + |O_i'|\right)^2},
\end{equation} 
where $M_i' = M_i-\bar{O}$ and $O_i'=O_i-\bar{O}$ and $\bar{O}$ is the mean of the observations. This score is bounded by zero and one and values close to one indicate good model skill. Since the Willmott skill score is nondimensional, it is an appropriate statistic to report when comparing skills of different variables, in this case, the total water level and the residuals. Since the total water level has a larger variability than the residuals the $\gamma^2$ measure should be much lower for the total water levels, hence, the Willmott skill score provides an additional point of comparison. These three statistics, along with the mean observed values and mean modelled values (which indicate the model's bias) are reported in Table \ref{tab:statistics} for each hindcast.

@@

%Local winds
The contribution to the surge due to atmospheric forcing is small and appears to be geographically dependent, most likely due to the effects of the local winds. Removal of the atmospheric forcing leads to a drop of about 2.5 cm in the maximum surge at Point Atkinson and 2.3 cm at Campbell River. By contrast, the surge amplitude increases by 4.8 cm at Victoria and 3.6 cm at Patricia Bay when atmospheric forces are neglected. The southeasterly winds over the Strait of Georgia could be acting to push water away from the southern tip of Vancouver Island, resulting in a slight decrease in the surge amplitude at Victoria and Patricia Bay when wind effects are included in the model. A simulation with only wind forcing produces a surge of 7.8~cm at Point Atkinson, a small fraction of the total observed surge, and only centimetres at the other locations.  Note that this wind-induced surge occurs well after the peak surge occurred in both the observations and simulations with full forcing.

@@

%local pressure
Next, the surge amplitude decreases by a few centimetres at all four locations if local atmospheric pressure is not included in the surface forcing. Other findings have reported the inverse barometer effect due to atmospheric pressure contributes up to $^1/_3$ of the total surge amplitude \citep{murty1995storm}. Note that in these simulations the removal of the atmospheric pressure in the surface forcing does not entirely remove the inverse barometer effect from the model results since the sea surface height anomaly at the open boundary includes an inverse barometer component. As such, these simulations indicate the effect of the local atmospheric pressure forcing but not the inverse barometer entirely. The contribution due to the inverse barometer can be calculated with equation (\ref{eq:inverse}) or by the approximation that 1 hPa drop in pressure corresponds to 1 cm rise in sea level. In this event, the pressure at the Vancouver International Airport dropped to about 99 000 Pa \citep{ECClimateArchive}, corresponding to a 20 cm rise in sea level. This inverse barometer contribution is more in line with the findings of \citep{murty1995storm} but does not alter the conclusion that the remote forcing contributes most significantly. Further, the model calculates the local atmospheric pressure forcing using a spatial gradient, and since the spatial scale of the low pressure system is large compared to the domain the local atmospheric pressure should not contribute greatly.

@@

Next, an assessment of the importance of including the tide-surge interaction, which is important in shallow regions due to nonlinear effects and bottom friction \citep{bernier2007tide}, is presented for this hindcast (Figure \ref{fig:tidesurge}). The most significant differences occur at Point Atkinson where the maximum of the surge-only case is 8.7 cm higher than maximum modelled residual, about 10\% of the total surge amplitude.  In contrast, the tide-surge interaction at Victoria is hardly noticeable. Indeed, a spatial map of the difference between the modelled residual and surge-only simulation (Figure \ref{fig:tidesurge_spatial}) indicates that the tide-surge interaction is more pronounced north of the San Juan and Gulf Islands. The passages between these islands are known for vigorous tidal mixing, which may have a nonlinear effect on the surge propagation. However, the effect is a relatively small overestimate of the surge elevation and no changes to the timing of the surge. So, in operational settings, performing a surge-only simulation and adding the predicted tides \textit{a posteriori} could be more computationally efficient with only a small decrease in the accuracy of the prediction.

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/feb2006_tidesurge_spatial.pdf}
\caption{Spatial dependency of tide-surge interaction in Feb 2006 hindcast. Difference in sea surface height between simulations with tidal forcing and without tidal forcing. Each plot is an average over 4 hours. }
\label{fig:tidesurge_spatial}
\end{figure}

@@

Modelled residuals are generally too low in this hindcast (Figure \ref{fig:dec2006}), however, the timing of the surge agrees well. All Willmott skill scores rank above 0.88 and the $\gamma^2$ measure also indicates good agreement between model and observations (Table \ref{tab:statistics}). The $\gamma^2$ measure suggests that the standard deviation of the error in the modelled residuals is roughly half of the standard deviation of the observed residuals. Note that the performance has degraded when compared to the Feb 2006 and Nov 2006 cases. This could be due to a number of factors. First, the very large wind speeds observed during this storm are significantly underrepresented by the atmospheric model leading to lower wind stress and less piling of water against the coast. Second, the anomaly forcing at the open boundaries could be too low. The observed residual at Tofino, BC (a point near by but outside of the domain) is up to 10 cm lower than the observed residual at Neah Bay, WA (a point near by but inside of the domain). Since the storm surge is less sensitive to wind forcing at the surface than anomaly forcing at the open boundaries, the latter is likely the cause.

@@

Hindcasts with both atmospheric forcing products agree well with the observations (Figure \ref{fig:dec2012_surge}). Willmott skill scores for both modelled residuals are above 0.92 at these four locations. Further, the modelled residuals deviate only slightly from each other, most noticeable at Point Atkinson during the large wind event on Dec 15, 2012 where simulation with the low resolution atmospheric forcing has a residual that is 5 cm higher. Although the wind speeds on Dec 15, 2012 are quite high, this event did not result in any flooding or residuals greater than 50 cm. Significant flooding occurred later, on Dec 17, 2012, where the observed surge amplitude at Point Atkinson was 68.3 cm. Even though the low resolution model significantly under predicts the wind speed at some locations, it performs comparably to forcing with high resolution products. Much of the fine scale detail offered by the high resolution atmospheric model is averaged out since, to first order, the storm surge is a spatial integral of the wind. This result further demonstrates that the local wind is not the most important factor for setting up storm surges over this domain.

@@

\section{Conclusion and discussion}\label{sec:diss}
In this article, we have configured a three-dimensional numerical ocean model of the Salish Sea for accurate hindcasts of storm surges in the Strait of Georgia. The model is forced with eight tidal constituents at the western and northern open boundaries, as well as sea surface height anomaly from de-tided observations at Tofino and Port Hardy. Additionally, wind stress and local atmospheric pressure forcing were used at the ocean surface using an atmospheric reforecasting product called CGRF \citep{smith2014new}. Storm surge simulations began with zero initial conditions for the currents and sea surface height and were commenced several days prior to the event of interest to spin up the velocity fields.

@@

Several factors considered in other storm surge modelling studies have been neglected in these simulations. Primarily, effects due to wetting and drying have not been included. This is an important factor to consider in shallow, low lying areas where water levels retreat significantly to reveal land. In this article, the four locations used to assess storm surge accuracy are not in shallow regions and so wetting and drying is not considered to be important. However, several communities in the Strait of Georgia, such as Boundary Bay and Crescent Beach, are in shallow regions and are typically most affected by storm surge during the winter season. As such, efforts to include wetting and drying will be considered in the future. Additionally, flooding effects due to wave run up is left for a more detailed inundation study with higher resolution in the near-shore areas. Local winds have a significant impact on wave action which is a primary concern for flooding management. Further, the remote forcing is applied uniformly across the open boundary and so any spatial variability in the incoming surge are not accounted for in the model. Finally, stratification effects on storm surge will be considered in a future study to determine if a barotropic model is reasonably accurate.

@@

Finally, difficulties in reproducing accurate M$_2$ tidal harmonics in both the Strait of Georgia and the Strait of Juan de Fuca have been highlighted in this study among others \citep{stronach1993update, foreman2004m}. Future work will attempt to improve model accuracy by examining mixing parametrizations and energy dissipation due to bottom friction and their effect on tidal harmonics. Additionally, a biological-chemical modelling component will be added.

@@

\appendices
\section{}\label{sec:appendix}
This section details the procedure used to tune the tidal forcing for a good match with observations in the Strait of Georgia. Originally, the eight most important tidal constituents were extracted from a tide prediction tool called Webtide \citep{webtide} at the western Juan de Fuca boundary based on work by \citet{foreman2000webtide}. At the northern boundary,  M$_2$ and K$_1$ tidal harmonics for both elevation and currents were taken from \citet{thomson1980johnstone}, as well as elevation harmonics for O$_1$ and S$_2$. The remaining constituents were extrapolated from Webtide based on the ratio between S$_2$/M$_2$ amplitudes and phase differences for the semi-diurnal constituents and O$_1$/K$_1$ for the diurnal constituents. However, these forcing conditions did not result in a good match between observed and modelled tidal elevation harmonics in the Strait of Georgia, the main region of interest (although the matches in the Strait of Juan de Fuca were adequate). As such, tidal forcing was tuned, by first adjusting the western boundary M$_2$ and K$_1$ phase and K$_1$ amplitude for a good match in the Strait of Georgia (specifically at Point Atkinson, Gibsons Landing, Winchelsea and Halfmoon Bay). This tuning was performed on simulations that used only M$_2$ and K$_1$ forcing. Next, the M$_2$ and K$_1$ fluxes at the northern boundary were adjusted because the cross-sectional area of transect used by \citet{thomson1980johnstone} to measure tidal currents in Johnstone Strait is smaller than the cross-sectional area of our northern boundary. Finally, the remaining constituents were added, starting with O$_1$, then S$_2$, and then in pairs P$_1$/N$_2$ and Q$_1$/K$_2$. Final adjustments were made by slightly varying amplitude and phase, evaluating the changes to M$_2$, K$_1$ and the ratios S$_2$/M$_2$, N$_2$/M$_2$, K$_2$/M$_2$, O$_1$/K$_1$, P$_1$/K$_1$, Q$_1$/K$_1$ and then choosing new values to move closer to observations by linear interpolation.
